Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-04-03 analysis/paper/paper.Rmd
Writing some more. Not limiting y-axis in WABI plot

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories -- for example as related to the production of blades or axes [@damlien2016]. Some studies have attempted more comprehensive evaluations of a number of lithic inventories to get at potential site types [@viken2018; @breivik2020], but these have been conducted without support from quantitative frameworks that can help structure such analyses, and have treated site type as a categorical variable with three categories. In sum therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods. The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites. This is done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency, and group sizes at visits to the sites. Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices will also influence how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In this light, the following sections will present the data to be analysed, a few potential sources of bias that might affect the comparisons to be conducted, while also highlighting a few ways in which idiosyncrasies in the chosen material might alleviate some of these issues.

@@

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. While the outer end of the first dimensions is dominated by later Mesolithic sites such as Stokke/Polland 3, Vallermyrene 4 and Gunnarsrød 10, which are all associated with axe production in non-flint materials, the later Mesolithic sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research.

In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite. The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2. T

@@

ggplot(lithics2, aes(log10(lithic_dens), log10(secondary_freq), 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(x = "Lithic volumetric density (log10)", 
       y = "Relative frequency of secondarily worked lithics (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme_classic()

@@

Figure \@ref(fig:wabi) displays the WABI as found for the site data. This indicates a negative correlation between the two variables (pearsons r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.

@@

The consistent nature of the WABI between this study (r = -0.50), @clark2017 (r = -0.50), @Bicho2020 (r = -0.48) is striking, and should speak to the cross-cultural and cross-temporal applicability of the WABI measure for lithic assemblages stemming from hunter-gatherer(-fisher) societies, and the potential it holds for comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, while this can perhaps support the of the . The question then becomes what is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, another possibility is that these are indicative of site that are temporally ubiquitous. In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020. This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to elements, and failed to identify any diachronic variation in this pattern. As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.

>>>>

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories -- for example as related to the production of blades or axes [@damlien2016]. Some studies have attempted more comprehensive evaluations of a number of lithic inventories to get at potential site types and associated mobility patterns [@viken2018; @breivik2020], but these have been conducted without support from quantitative frameworks that can help structure such analyses. In sum therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods. The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites. This is done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency, and group sizes at visits to the sites. Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices will also influence how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In this light, the following sections will present the data to be analysed, a few potential sources of bias that might affect the comparisons to be conducted, while also highlighting a few ways in which idiosyncrasies in the chosen material might alleviate some of these issues.

@@

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research.

In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite. The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2.

# Hunter-gatherer mobility and whole assemblage behavioral indicators

The results of the correspondence analysis does appear to align reasonably well with previous research that is based on more informal analyses of, indicating that the employed artefact categories are behavourally meaningful. Aligning the signficance of this chronological patterning to specific behavioural dimensions, is, however, another task entirely, and a task that is severely hampered by fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, and consequently how the different debitage and tool categories should be treated @dibble2017. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tools could have had various different purposes and had a multitude of shapes throughout their use-life. Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis and an overly naive trust in the fact that these units reflect

@@

ggplot(lithics2, aes(log10(lithic_dens), log10(secondary_freq * 100), 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(x = "Lithic volumetric density (log10)", 
       y = "Relative frequency of secondarily worked lithics (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme_classic()

@@

Figure \@ref(fig:wabi) displays the WABI as found for the site data. This indicates a negative correlation between the two variables (pearsons r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.

@@

The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking, speaks to the cross-cultural and cross-temporal applicability of the measure, the potential it holds for comparative studies, and is to be expected given the . The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, while this can perhaps support the of the . The question then becomes what is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, another possibility is that these are indicative of site that are temporally ubiquitous. In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020. This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to elements, and failed to identify any diachronic variation in this pattern. As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.
